Multi-level modeling with Openflexo/\FML

### 2021-10-15 responseLetter.tex

\begin{response}{The reviewers have made suggestions for improving the presentation but also raised a number of issues requiring clarification. An overarching theme is the request for a more explicit description of how the multi-level modelling emulation is achieved and what its general advantages and limitations are, including a clarification of what the implications would be of supporting more "levels" than the challenge required.}
%We have been more explicit to explain how we use the FML language at the end of the introduction, and in various figures. TODO.

We have improved the presentation of both, Openflexo/FML and our solution to the challenge so that: 1) the techniques we use to \emph{achieve} multi-level capabilities are more explicit; and, 2) the advantages and disadvantages of using them in order to fulfil the challenge requirements become clearer. Notably, we have added many details in the final paragraphs of the introduction section, improved the Section 2, devoted to the description of Openflexo, refactored figures 2, 4, 5, 6, 7 and 9 and, added figures 3 and 10.
\end{response}

\begin{response}{We very much welcome a "conventional" response to the challenge but would like to see the respective reviewer questions addressed in a revised version which is explicit as possible about the advantages and the disadvantages of the emulation. You may find the references listed below useful to keep explanations concise by referencing related work (note that reviewer 3's "Type-Instance pattern" is also known as the "Type Object" pattern). You should also respond to other concerns reviewers raise, either by making changes to the submission or in the form of an authors' response.}

\textcolor{orange}{In the answer to the previous editor's comment we describe the improvements of our revision, which includes an emphasis in discussing trade-offs related to the emulation of multi-level modeling by using the Openflexo/FML and the type-object pattern. Additionally, the related work section has been extended in order to enhance the comparison between our approach and previous solutions to the challenge.}
% fd ok
% sg ok
% jc ok

@@

\Reviewer{\#1}
\begin{response}{In section 2 Technology the authors describe the foundations of the Openflexo approach. Here it would greatly help the reader to see an example how the approach works for traditional modeling languages. For example, how would you realize a small subset of BPMN with the approach? This could briefly be illustrated so as to give the reader an immediate understanding. Especially, the 'metamodel'-agnostic approach seems very interesting but I would like to see how that is used for traditional modeling approaches.}

The Figure \ref{fig:BPMNSubsetExample} (above) shows how to realize a small subset of BPMN with Openflexo/FML (assuming a metamodel of the full specification of BPMN exists). Concretely, we show  how a virtual model (\texttt{SimplifiedBPMN}) can filter an existing model (BPMN) and be instantiated. Note that connection with external sources is not required for our solution to the challenge and is only represented in order to illustrate the capabilities of Openflexo/FML. Therefore, in the article we choose not to give this level of details and focus on the left part of the figure.
% Fabien OK pour moi
% sg ok
% jc ok

\begin{figure}[t]
    \centering
    \includegraphics[width=1.0 \textwidth]{Figures/BPMNSubsetExampleWithExternalConnexion.pdf}
    \caption{An example of BPMN subset built on top of a BPMN/EMF model}
    \label{fig:BPMNSubsetExample}
\end{figure}

\todo{SG: Est ce possible de rajouter ce listing ou ca n'avance à rien ?}

@@

%\begin{lstlisting}
%use org.openflexo.technologyadapter.emf.EMFModelSlot as EMF;
%import ["http://www.omg.org/spec/BPMN/20100524/MODEL-XMI"] as BPMN_METAMODEL;
%
%// A simplified BPMN metamodel exposing
%// both concepts BPMNProcess and BPMNTask
%model SimplifiedBPMN {
%
%  // Connexion to a .bpmn file
%  EMFModel bpmnModel
%    with EMF::EMFModelSlot(
%        metaModel = BPMN_METAMODEL);
%
%  // Abstract concept BPMNElement defining
%  // a name and a link to an eventual BPMNElement
%  // in .bpmn file
%  abstract concept BPMNElement {
%  	EMFObjectIndividual bpmnElement
%  	    with EMF::EMFObjectIndividualRole (
%			source = bpmnModel
%		);
%    String name values bpmnElement.name;
%  }
%
%  concept BPMNProcess extends BPMNElement {
%
%    // Describe here properties and
%    // behaviours for BPMNProcess concept
%    ...
%
%    concept BPMNTask extends BPMNElement {
%
%        // Describe here properties and
%        // behaviours for BPMNTask concept
%        ...
%
%    }
%
%    // Other core concepts
%  }
%}
%\textbf{}\end{lstlisting}

\end{response}

\begin{response}{Also, at the stage of section 2, the characterization of the FML language should be made clearer. The metamodel in Figure 2 seems mainly conceptual - but at the same time you refer to it as a basis for implementation. Could that be clarified? eg you have the class "Behavior" - which is a concept and I did not understand right away what it is composed of. On the other hand you have for example the class "ModelSlot" which seems like a very technical realization. Also, you note that roles have types but I could not find any types in the metamodel (?). Thus, I propose to include a more detailed version of your metamodel at this stage. Please note that EMISAJ does not have a page limitation, so no worries if that should require more space.}

\textcolor{orange}{We agree with the reviewer. The previous version of the figure depicting the main concepts of the FML language (Figure 2) was too succinct. We have fully refactored it so that now important concepts such as \textsf{Type} are shown. We also have correspondingly improved the text explaining the figure. We hope this changes make understanding the core concepts of the FML language easier.}
%fd ok, j'ajouterais bien aussi :
{\color{teal} Furthermore, to make it easier to distinguish the concepts of FML, the figure 2 adopts a slightly different graphical syntax while the caption insists on the fact that it is a representation. Indeed, FML is implemented by an interpreter that understands these concepts. For example, defining a behavior consists in providing its code using the FML language syntax (like in a programming language).}
% sg d'accord avec cette proposition
% jc ok

@@

\pagebreak

\Reviewer{\#2}

\begin{response}{The basic idea of model federation is interesting, its primary focus is not on creating multi-level models. However, as the approach is quite flexible, multi-level behavior can be emulated by model federation. This fact is possibly the most serious advantage and disadvantage of the paper at the same time. Advantage, since it brings us a completely new, different way of thinking, and disadvantage since it does not really help in categorizing/unifying multi-level methods and -- for me -- it also feels a little bit off-topic in this special issue. At this point, it should be mentioned that the authors managed to cover/solve all requirements of the challenge, thus the paper, as a solution to the challenge fits well.}

\textcolor{orange}{We fully agree with the reviewer on the fact that it is the flexibility of the model federation approach that enables us to provide a solution to the multi-level process challenge while using a tool not specifically designed for multi-level modelling. Of course, this comes together with a number of trade-offs which we discuss in the paper (\emph{e.g.}, see the the end of the introduction section).}
%fd ok
% sg ok
\textcolor{red}{We have included at the end of the introduction section a more detailed description of our approach. In a word, we use FML (linguistic level) as a tool to build models and meta-models permitting us to emulate multi-level modelling. The evolution of models (multi-levels) is encoded as a hierarchy of models and meta-models. In this sense, we do on-demand multi-level modelling, and propose an organization guided by the evolution of models.}
% jc ok

\textcolor{teal}{Propale de Fabien pour compléter la réponse ici. We have added a discussion (où?) on the fact that our approach reifies the levels using "generic" models and relies on the process of construction to implement the levels of the challenge. We hope, it makes it easier to see that it is a contribution to multi-level domain.}

% sg ok : je pense que c'est dans la section 3
% jc euh… je ne vois pas dans la section 3.

@@

\begin{response}{The third section contains an overview of the solution. Here, virtual models are instantiated to virtual model instances conforming to the virtual models acting as the metamodel for the virtual model instances. The question naturally arises: how does it work, if we have more abstraction levels?  In this section, the authors also discuss linguistic and ontologic instantiation. The explanation refers to Section 4 resulting in a hardly understandable analysis of different instantiation methods. Figure 4. does not help in clarifying the concepts either (e.g. using the name InstanceInstance seems to be a very bad decision). From the point of the muli-level behavior of view, the interpretation of the instantiation relationship is the most crucial one that is why this section must be clarified. Where are the rules of ontological instantiation defined, how are they enforced? The text mentions the (linguistic) instantiation mechanism of the FML Language… how does it work exactly?}

\textcolor{orange}{As indicated in the answer to the previous comment we have refactored Section 2 in order to provide more details and make explanations easier to follow. We have done the same with Section 3. Note that \texttt{InstanceInstance} has been replaced with \texttt{ObjectInstance}} \todo{Probably it would be nice to add here some concrete answers instead of directing the reviewer to read two sections.}

{\color{red} FD: je pense aussi qu'il faut être plus précis dans cette réponse. J'aurais bien dit que l'instantiation linguistique consiste juste à utiliser le langage FML et respecter sa syntaxe et ses contraintes de typage et que pour l'ontologique, c'est uniquement par le process ou l'on s'impose de faire des factory depuis le type (Sylvain ?)}

{\color{red} Sylvain: (Nouvelle proposition de réponse) As indicated in the answer to the previous comment we have refactored Section 2 and 3 in order to provide more details and make explanations easier to follow. Linguistic instantiation in this context refers to the use of FML language for our solution. This offers syntax and semantics, as well as type checking and various constraints checking. We tried to show how the use of FML helps in meeting requirements of the challenge. On the other side, we exploit the ontologic instantiation through an extensive use of the "Type Object" pattern in two versions (with single and multiple cardinalities), controlling the ontologic instantiation from the type, using a classical factory pattern. Additionaly, we agree that the choice of \texttt{InstanceInstance} was not a good decision, we have replaced it with \texttt{ObjectInstance}.}
% jc ok (avec la nouvelle proposition de Sylvain)

@@

% Fabien ok pour moi

\begin{response}{Overall, the solutions for the requirements are correctly presented in this section, but there are some spots not clear enough. E.g. regarding ontological instantiation, the solution uses two kinds: single and multiple. More details would be welcome to see, what their exact semantics are and where do we enforce these semantic rules (is it enforced by the engine? Is the semantic hard-coded?). What are the contexts of the constraints (to which concept are they attached to)?}
MultiInstance(s) are introduced in our answer metamodel (figure 6). As such, their semantic is defined at the ontological level, i.e. in the metamodel. It is an ad hoc mechanism introduced to fulfill requirements.

{\color{red}fd: hum, j'ai l'impression qu'il manque un peu de détail pour répondre sur la partie multi-instance. Il me semble qu'en plus d'être ad hoc, le mécanisme est manuel (mais je me trompe peut être). Si c'est vrai, il faudrait le dire.}

{\color{red}Sylvain: (Nouvelle proposition de réponse) MultiInstance(s) are introduced in our answer metamodel (figure 6). As such, their semantic is defined at the ontological level, i.e. in the metamodel. It is an ad hoc mechanism introduced to fulfill requirements. This means that corresponding semantics is not handled by FML but in the designed virtual models, with specific behaviours implementing underlying semantics.}

@@

The constraints are defined at FlexoConcept level, and are designed of invariant being evaluated by the FML interpreter, which may raise constraint violation at run-time. An example of constraint is presented for requirement P9, in \texttt{TaskType} concept (p 8).

% jc ok
\end{response}

\begin{response}{In Section 4.1.2 a behavior called newProcess is defined. Is it called automatically (by the framework) when creating a new process, or should the users call it? Is this a convention, or are such methods configured somehow to replace the usual constructors? According to the text, "same pattern applies" in several cases, but these patterns could be also included, because it is not completely clear to what part of the behavior "same" refers. In same cases, the diagram uses "/type", while in the code it is "type" -- why? The specialization of Actors is encoded by FML, wouldn’t be easier to use the built-in specialization instead?}

{\color{teal}The behavior \texttt{newProcess} is defined in the \texttt{ProcessType} concept following a \emph{factory} design pattern. Rather than being created directly a process instance is therefore created from its type. This realizes the ontological instantiation. There is no notion "usual" or "default" constructors in FML. Whenever a user creates a new instance, it has to chose one constructor. Here, we impose methodologically to the user to create a process instance through the chosen process type. There is no enforcement in our solution. SYLVAIN ?

Concerning the use of "/type" and "type", the former one is a way to stress that the relation is derived, i.e. it is a refinement of the "type" relation defined in another level. In the FML code, we use "type" due to the fact there is no notion of level. (It is explained in the lasta paragraph of 4.1 on page 8.)}
\textcolor{red}{
%- newProcess est un vrai behaviour défini dans ProcessType (pattern factory) et permet de spécifier l'instantiation d'un Process à partir d'un ProcessType (c'est la partie réalisation de l'instantiation ontologique)
\begin{itemize}
\item difficile de répondre sur le "same pattern applies"  (fd finalement, je pense qu'il ne comprend pas bien notre usage de "same pattern" car il a l'impression qu'il y a plusieurs pattern => peut etre faudrait-il les nommer (specialization, factory, ...)) ; (jc) j'ai l'impression que la question est un peu à côté de la plaque
\item pas compris la question sur la spécialisation des Actors (fd ne comprend pas la question, jc non plus)
\end{itemize}
}

@@

\end{response}

\begin{response}{In Section 4.2 it is mentioned, where the solution uses linguistic instantiation, but it is not clear by which component it is handled and what rules are followed when instantiating the concepts. In Section 4.3, tooling is elaborated. I suggest removing redundant information about the two editors and add more details on how the concrete syntax (e.g. the texts in the box of Fig. 9.) is defined.}
We have improved the description of the relationship between FML and our metamodels (use of colors in many figures).
The tooling uses model federation, but the description of our answer relies only on \textsf{VirtualModel} and \textsf{FlexoConcept} of FML (figure 2). The tooling is suggested in figure (new) 9. The tooling is not detailed, but the virtual model \texttt{ProcessEditor} presented in figure (new) 9 defines a lot of interactions, where texts of (old) figure 9 are found.

The business logic of the diagram tooling is encoded in FML, and more specifically for this editor in the \texttt{ProcessEditor} virtual model presented in figure 9. Structure of diagramming, interactions and presentation layer (\emph{e.g.} texts) are defined in various concepts defined in the virtual model \texttt{ProcessEditor}.
%fd ok
%sg ok
%jc ok
\end{response}

\begin{response}{In Section 5, the meaning of the category Conceptualization is hard to understand, does this mean that a concept is created that covers the requirement completely? In the case of ontological instantiation, the term “dual concept nature” and “instance nature” is mentioned… what do these terms mean?}

%On a rectifié et ré-écrit ces bouts de texte
%TODO (Salvador).

\textcolor{orange}{Conceptualization means that a concept is created to fulfill or meet a requirement. As for the mention to a "dual concept nature", we have rewritten that part as it was indeed confusing. Now it reads: the fulfillment of a requirement is obtained by the relation between an instance and its type definition instance.} 
%jc ok

@@

\begin{response}{Section 7 contains a rather short elaboration on related approaches. Although I agree that it is enough to focus on previous solutions here, much more detail and a comparison would be required.}

We have extended Section 7. It discusses now previously published solutions to the challenge in more detail. Comparisons with our approach have been also extended.

\end{response}

\begin{response}{As Section 6 shows, the authors know exactly, why their solution may have difficulties as a multi-level modeling approach despite solving the given challenge. The solution is basically a smart, two-level modeling solution of the MULTI challenge. Multi-level behavior is emulated by constraints and additional mechanisms particularly written to this challenge. The result is a working workbench fulfilling the requirements but having limited reusability and extensibility. For example, it would be hard to increase the number of abstraction levels as class-like and object-like concepts are handled differently. Another problem may arise if we would use the framework as a prototyping tool and change the metamodel and the models at the same time.}

{\color{teal}hum, je crois qu'il n'a pas bien compris mais c'est difficile de le dire comme cela. je pense que l'on pourrait indiquer que chaque niveau est représenté par un virtual model et donc ajouter un niveau se ferait en ajoutant un virtual model. En revanche, il y a peu de support automatisé et nous reposons sur une approche manuel pour gérer correctement ces niveaux. La limite est probablement plus dans la faible aide (me semble t il) Et pour le dernier point, comme notre outil repose sur de l'interpretation il est possible de travailler au deux niveaux à la fois. Par contre, il faut une certaine discipline pour ne pas tout casser (SYLVAIN ?)}

{\color{red}Sylvain: (Nouvelle proposition de réponse) Indeed, our solution relies on a ad-hoc architecture where multi-level is emulated. However, our solution presents reified layers though virtual model inheritance hierarchy. Although there is no automatized support, the proposed methodology allows to provide an arbitrary number of layers and some extension capabilities and reuse support (e.g. adding a level would mean creating a new virtual model extending an existing one). This is made possible by the fact that FML allows one to work on different levels at a time. The limitations of this approach resides in a good comprehension of the architecture and a strong expertise to maintain global consistency.}
%jc ok (j'ai probablement loupé la première réponse)
\end{response}

\begin{response}{Minor issues/questions:
\begin{itemize}
    \item Are there any tools supporting model federation besides Openflexo? \textcolor{orange}{To the best of our knowledge there are not other model federation tools comparable to Openflexo/FML.}
    \item Section 2. "In contrast to approaches that compose metamodels into a single large metamodel grouping all needed entities,…" - I do not see, why this is relevant here \textcolor{orange}{Our intention was to compare the federation approach to approaches that integrate all metamodels together. We argue that reifying models (with virtual models) helps structuring, organizing, federating models.}
    \item Figure 5 - concept Sequencing - the roles "in" and "out" are not represented by lines… why? \textcolor{orange}{As explained in a previous answer, we do this for the sake of readability.}
    \item What does the term "sub-concept" mean exactly? Please explain it before the first use \textcolor{orange}{Indeed, this was unclear. This was related to concept inheritance. We have removed its use.}
    \item ProgrammingLanguage concept is an enumeration… how is it modelled? \textcolor{orange}{ProgrammingLanguages is modelled as an enumeration of names (Java, Cobol, etc) using the notion of enumeration of the FML language. }
    \item "The Acme metamodel also defines a specialized task type Coding" - isn’t it CodingTask? \textcolor{orange}{Yes, this is correct. Coding is actually a CodingTask, as shown in figure 7.}
    \item "The value of the property producedArtifactTypes" - is it property, or role? \textcolor{orange}{Indeed, this was unclear. Figure 2 has been completed to show all kind of properties (\texttt{FlexoRole} is a specialization of \texttt{FlexoProperty}). The two terms "property" and "role" are equivalent in the context of \texttt{FlexoRole} use. }
    % fd: TODO, Syl-vain, il me semble que c’est la seule occurrence de property,on pourrait basculer sur role il me semble
\end{itemize}}

### 2021-10-15 responseLetter.tex

\begin{response}{(3) One of the key benefits of the Type-Instance pattern is that it provides a 'hook' on which semantics can be defined using constraints. There is a reified relationships between a type and its instance and therefore conditions can be defined that establish both static and dynamic semantics. I am confused regarding why this is not a key feature of the semantics for the challenge as it is implemented by the authors. Constraints are provided very sparingly compared to implementation of behaviour in FML. Unfortunately, since the dynamic semantics of FML is not provided (and is bespoke) this is less clear than it might otherwise be for the reader. Perhaps it would be better to use constraints in the context of the Type-Instance pattern to specify semantics? It would be more abstract than the current 'method' style semantics.}
{\color{red}
We designed our answer with the Type/Instance pattern. We think, we could have implemented it with another approach such as Clabject.
We had to define the expected semantic of the challenge. To do that, FML offers constraints, but we can also define our own constraint at the ontological level (metamodel). We have here a trade-off. Reusing the constraint from FML is efficient but may be too restrictive for the user. Defining our own constraints is more adaptable, but it has to be implemented and the produced model may be incorrect. Due to the size of the problem and the expected freedom, we chose the second way. To make it easier to build correct models, we propose several built-in behavior in charge of the various operation and notably the instantiation operations. This result in a "constrained" tool rather than a "restrictive" metamodel.

Our implementation of the challenge was also an opportunity to showcase the expressivity of FML. So we tried, in our solution, to reuse several mechanism of FML. For example, sometime we use behavior ({\color{teal} P17, P18 and P19}) and sometimes we use constraints ({\color{teal} P9 and P17}).}

### 2021-10-18 responseLetter.tex

\Reviewer{\#1}
\begin{response}{In section 2 Technology the authors describe the foundations of the Openflexo approach. Here it would greatly help the reader to see an example how the approach works for traditional modeling languages. For example, how would you realize a small subset of BPMN with the approach? This could briefly be illustrated so as to give the reader an immediate understanding. Especially, the 'metamodel'-agnostic approach seems very interesting but I would like to see how that is used for traditional modeling approaches.}

The Figure \ref{fig:BPMNSubsetExample} (above) shows how to realize a small subset of BPMN with Openflexo/FML (assuming a metamodel of the full specification of BPMN exists). Concretely, we show  how a virtual model (\texttt{SimplifiedBPMN}) can filter an existing model (BPMN) and be instantiated. Note that connection with external sources is not required for our solution to the challenge and is only represented in order to illustrate the capabilities of Openflexo/FML. Therefore, in the article we choose not to give this level of details and focus on the left part of the figure.
% Fabien OK pour moi
% sg ok
% jc ok

\begin{figure}[t]
    \centering
    \includegraphics[width=1.0 \textwidth]{Figures/BPMNSubsetExampleWithExternalConnexion.pdf}
    \caption{An example of BPMN subset built on top of a BPMN/EMF model}
    \label{fig:BPMNSubsetExample}
\end{figure}

\todo{SG: Est ce possible de rajouter ce listing ou ca n'avance à rien ?. SM: If you want to add it, it should be ok, but remember then to update the answer text.}

@@

{\color{teal}The behavior \texttt{newProcess} is defined in the \texttt{ProcessType} concept following a \emph{factory} design pattern. Rather than being created directly a process instance is therefore created from its type. This realizes the ontological instantiation. There is no notion "usual" or "default" constructors in FML. Whenever a user creates a new instance, it has to chose one constructor. Here, we impose methodologically to the user to create a process instance through the chosen process type. There is no enforcement in our solution. SYLVAIN ?

Concerning the use of "/type" and "type", the former one is a way to stress that the relation is derived, i.e. it is a refinement of the "type" relation defined in another level. In the FML code, we use "type" due to the fact there is no notion of level. (It is explained in the lasta paragraph of 4.1 on page 8.)}
\textcolor{red}{
%- newProcess est un vrai behaviour défini dans ProcessType (pattern factory) et permet de spécifier l'instantiation d'un Process à partir d'un ProcessType (c'est la partie réalisation de l'instantiation ontologique)
\begin{itemize}
\item difficile de répondre sur le "same pattern applies"  (fd finalement, je pense qu'il ne comprend pas bien notre usage de "same pattern" car il a l'impression qu'il y a plusieurs pattern => peut etre faudrait-il les nommer (specialization, factory, ...)) ; (jc) j'ai l'impression que la question est un peu à côté de la plaque
\item pas compris la question sur la spécialisation des Actors (fd ne comprend pas la question, jc non plus)
\end{itemize}
}

\end{response}

\begin{response}{In Section 4.2 it is mentioned, where the solution uses linguistic instantiation, but it is not clear by which component it is handled and what rules are followed when instantiating the concepts. In Section 4.3, tooling is elaborated. I suggest removing redundant information about the two editors and add more details on how the concrete syntax (e.g. the texts in the box of Fig. 9.) is defined.}

We have improved the description of the relationship between FML and our metamodels (use of colors in many figures).
The tooling uses model federation, but the description of our answer relies only on \textsf{VirtualModel} and \textsf{FlexoConcept} of FML (figure 2). The tooling is suggested in figure (new) 9. The tooling is not detailed, but the virtual model \texttt{ProcessEditor} presented in figure (new) 9 defines a lot of interactions, where texts of (old) figure 9 are found.

The business logic of the diagram tooling is encoded in FML, and more specifically for this editor in the \texttt{ProcessEditor} virtual model presented in figure 9. Structure of diagramming, interactions and presentation layer (\emph{e.g.} texts) are defined in various concepts defined in the virtual model \texttt{ProcessEditor}.
%fd ok
%sg ok
%jc ok
%sm ok
\end{response}

@@

\end{response}

\begin{response}{As Section 6 shows, the authors know exactly, why their solution may have difficulties as a multi-level modeling approach despite solving the given challenge. The solution is basically a smart, two-level modeling solution of the MULTI challenge. Multi-level behavior is emulated by constraints and additional mechanisms particularly written to this challenge. The result is a working workbench fulfilling the requirements but having limited reusability and extensibility. For example, it would be hard to increase the number of abstraction levels as class-like and object-like concepts are handled differently. Another problem may arise if we would use the framework as a prototyping tool and change the metamodel and the models at the same time.}

{\color{teal}hum, je crois qu'il n'a pas bien compris mais c'est difficile de le dire comme cela. je pense que l'on pourrait indiquer que chaque niveau est représenté par un virtual model et donc ajouter un niveau se ferait en ajoutant un virtual model. En revanche, il y a peu de support automatisé et nous reposons sur une approche manuel pour gérer correctement ces niveaux. La limite est probablement plus dans la faible aide (me semble t il) Et pour le dernier point, comme notre outil repose sur de l'interpretation il est possible de travailler au deux niveaux à la fois. Par contre, il faut une certaine discipline pour ne pas tout casser (SYLVAIN ?)}

{\color{red}Sylvain: (Nouvelle proposition de réponse) Indeed, our solution relies on a ad-hoc architecture where multi-level is emulated. However, our solution presents reified layers though virtual model inheritance hierarchy. Although there is no automatized support, the proposed methodology allows to provide an arbitrary number of layers and some extension capabilities and reuse support (e.g. adding a level would mean creating a new virtual model extending an existing one). This is made possible by the fact that FML allows one to work on different levels at a time. The limitations of this approach resides in a good comprehension of the architecture and a strong expertise to maintain global consistency.}
%jc ok (j'ai probablement loupé la première réponse)
%sm ok.. but we could even ignore this comment as it does not really request action.
\end{response}

@@

%You are right. We intensively use the Type/Instance pattern.
%We made this choice because it was easier to implement with Openflexo.
%We make it more explicit in the text.
\end{response}

\begin{response}{(2) In section 1 the authors list 2 key reasons why a strict approach to modelling is sub-optimal. However, they do not establish the case that their approach uses these 2 features to any concrete benefit. It would be appropriate (perhaps at the end of the article) to revisit these features and to clearly demonstrate that their approach supports them and that they have concrete utility in some way in the context of the challenge.}

We have added a paragraph in the Conclusions section indicating how Openflexo/FML permitted us to overcome the mentioned limitations of the strict modeling paradigm and how this was used in order to build our solution to the Multi-Level Process challenge. Notably, we mention how the type/object pattern allowed us to use model element instances as types in order to achieve ontological instantiation.
%fd ok
%sg ok
\end{response}

\begin{response}{(3) One of the key benefits of the Type-Instance pattern is that it provides a 'hook' on which semantics can be defined using constraints. There is a reified relationships between a type and its instance and therefore conditions can be defined that establish both static and dynamic semantics. I am confused regarding why this is not a key feature of the semantics for the challenge as it is implemented by the authors. Constraints are provided very sparingly compared to implementation of behaviour in FML. Unfortunately, since the dynamic semantics of FML is not provided (and is bespoke) this is less clear than it might otherwise be for the reader. Perhaps it would be better to use constraints in the context of the Type-Instance pattern to specify semantics? It would be more abstract than the current 'method' style semantics.}
{\color{red}
We designed our answer with the Type/Instance pattern. We think, we could have implemented it with another approach such as Clabject.
We had to define the expected semantic of the challenge. To do that, FML offers constraints, but we can also define our own constraint at the ontological level (metamodel). We have here a trade-off. Reusing the constraint from FML is efficient but may be too restrictive for the user. Defining our own constraints is more adaptable, but it has to be implemented and the produced model may be incorrect. Due to the size of the problem and the expected freedom, we chose the second way. To make it easier to build correct models, we propose several built-in behavior in charge of the various operations and notably the instantiation operations. This result in a "constrained" tool rather than a "restrictive" metamodel.

@@

\end{response}

\begin{response}{(4) Section 6 lists a number of benefits of the approach. Unfortunately, the presentation in the section is very general and none of the claims are justified in terms of concrete descriptions and/or examples. This section seems to be very important and could be of great interest to the reader. Perhaps it would be appropriate to reduce some of the rather implementation heavy descriptions earlier in the paper and increase the analysis in this section? In addition section 4.3 notes that MLM is key to the ability of Openflexo to create tooling. Again, this is very interesting, but no concrete details are provided as to how Openflexo uses the level agnotic features of the models to create tools. Can the essence of this be captured for Openflexo and then described in terms of how it facilitiates the tooling arising from the challenge?}

\textcolor{orange}{We have refactored the subsections of Section 6 in order to include concrete examples and references to the description of our solution whenever it was possible and/or relevant. Concretely, subsections 6.2, 6.3, 6.4, 6.6, 6.7 and 6.11 have been updated.}

\todo[inline]{Here we have to explain how openflexo helps to create tooling. Just a line or two should suffice. Or maybe I have misanderstood the question... Is the reviewer really wrong and thought we were talking about MLM? Then just copy answer below from: Concerning...}

### 2021-10-18 3-introduction.tex
replace typo VirtalModel
synchro

This strict approach shows its limitations when modeling complex domains requiring more than one level of specialization, \eg to adapt the model to application sub domains. Indeed, the two-level approach fails to acknowledge and support: 1) the existence of different forms of classification and/or instantiation (\eg ontological vs linguistics); and 2) the duality type-object for model elements. Therefore, while it may still be used for the aforementioned complex modeling tasks, the strict approach entails the introduction of accidental complexity in both the modeling process and the resulting modeling artefacts.

@@

Our modeling approach is based on a language (\FML) which provides the (linguistics) concepts that are used to define  the multi-levels models and meta-models that are needed. \FML allows to define models as first class entities (which we call VirtualModel) and to link them together (by extension). In the \mpc case study, we exploit this possibility to:
\begin{itemize}
    \item Define, extend and adapt models, keeping the history, to meet the %changing
    modeling requirements of the case study,
    \item Align this hierarchy of models with requirements evolution ($P_1-P_{19}$ then $S_1-S_{13}$).
\end{itemize}
This methodological choice allows us to adapt the models and keep track of their evolution each time the requirements evolve.

Contrary to a classical multi-level approach, in which the notion of multi-level is integrated in the language and must therefore be respected, our approach allows to build multi-levels on demand, and with the appropriate form. Without multilevel-specific operators/concepts, the "custom" construction must be done explicitly during the model development and evolution process.

>>>>

This strict approach shows its limitations when modeling complex domains requiring more than one level of specialization, \eg to adapt the model to application subdomains. Indeed, the two-level approach fails to acknowledge and support: 1) the existence of different forms of classification and/or instantiation (\eg ontological vs linguistics); and 2) the duality type-object for model elements. Therefore, while it may still be used for the aforementioned complex modeling tasks, the strict approach entails the introduction of accidental complexity in both the modeling process and the resulting modeling artefacts.

@@

Our modeling approach is based on a language (\FML) which provides the (linguistics) concepts that are used to define the multi-level models and meta-models that are needed. \FML allows defining models as first class entities (which we call \emph{virtual models}) and to link them together (by extension). In the \mpc case study, we exploit this possibility to:
\begin{itemize}
    \item Define, extend and adapt models, keeping the history, to meet the %changing
    modeling requirements of the case study,
    \item Align this hierarchy of models with requirements evolution ($P_1-P_{19}$ then $S_1-S_{13}$).
\end{itemize}
This methodological choice allows us to adapt the models and keep track of their evolution each time the requirements evolve.

Contrary to a classical multi-level approach, in which the notion of multi-level is integrated in the language and must therefore be respected, our approach allows building multi-level on demand, and with the appropriate form. Without multilevel-specific operators/concepts, the "custom" construction must be done explicitly during the model development and evolution process.

### 2021-10-18 4-technology.tex
synchro

To illustrate this, figure~\ref{fig:BPMNSubsetExample} shows how to realize a small subset of BPMN with Openflexo/FML. Concretely, we have designed a \textit{virtual model} (\texttt{SimplifiedBPMN}) with 3 \textit{concepts} representing a process and its contained tasks. An instance of this \textit{virtual model} is presented in the bottom of the figure, showing instances with their \textsf{instanceOf} relationships with their model.
%
The following listing shows the \FML code of this \textit{virtual model}. {\color{red}Notice that the composition relation of figure~\ref{fig:BPMNSubsetExample} between \texttt{BPMNProcess} and \texttt{BPMNTask} is realized in the FML code by concept containment. Whenever an instance of \texttt{BPMNTask} is created, it must be in a container (instance of \texttt{BPMNProcess}) and the two instances get linked together.}\noteFabien{A valider}

>>>>

To illustrate this, figure~\ref{fig:BPMNSubsetExample} shows how to realize a small subset of BPMN with Openflexo/FML. Concretely, we have designed a \textit{virtual model} (\texttt{SimplifiedBPMN}) with 3 \textit{concepts} representing a process and its contained tasks. An instance of this \textit{virtual model} is presented in the bottom of the figure, showing instances with their \textsf{instanceOf} relationships with their model.
%
The following listing shows the \FML code of this \textit{virtual model}. Note that the composition relation of figure~\ref{fig:BPMNSubsetExample} between \texttt{BPMNProcess} and \texttt{BPMNTask} is realized in the FML code by concept containment. Whenever an instance of \texttt{BPMNTask} is created, it must be in a container (instance of \texttt{BPMNProcess}) and the two instances get linked together. %\noteFabien{A valider}
